Frequency of exposure mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-02-26 manuscript/manuscript.qmd
Add appendices

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition"
editor: source
author:
  - name: Gonzalo Garcia-Castro
    orcid: 0000-0002-8553-4209
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
    corresponding: true
  - name: Daniela S. Ávila-Varela
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra
  - name: Ignacio Castillejo
    affiliations:
      - Departamento de Psicología, Universidad Autónoma de Madrid
  - name: Núria Sebastian-Galles
    affiliations:
      - Center for Brain and Cognition, Universitat Pompeu Fabra

affiliation: Center for Brain and Cognition, Universitat Pompeu Fabra

@@

abstract: |
    Recent studies have reported larger vocabulary sizes for bilingual toddlers learning languages that shared a greater amount of cognates (e.g., English-Dutch). This effect was only observed in the language to which toddlers were exposed the least, and has been suggested to be the result of a cognateness facilitation effect during word acquisition. The specific underpinnings of tyhis effect, and its relationship with the degree of exposure to the language are unclear. In this study, we present vocabulary data from 436 Catalan-Spanish bilinguals and monolinguals aged 12 to 34. We used Bayesian Item Response Theory to estimate participants' probability of acquisition of 604 words, conditional to their word's cognate status and lexical frequency, and the child's age and degree of exposure to each language. We found converging evidence for an earlier age of acquisition for cognate words, mediated by word frequeny and language exposure. Low-frequency words, and words from the language of least exposure benefitted more strongly by their cognate status than high-frequency words. Our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition and processing of words in the other language.

@@

keywords: lexical acquisition, vocabulary, bilingualism, item response theory, bayesian
toc: false
citation: true
fig-dpi: 1000

csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"

format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(ggplot2)
library(patchwork)
library(ggtext)
library(lubridate)
library(brms)
library(glue)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(scales)
library(bvqdev)

@@

tar_load(model_rhats)
tar_load(model_neffs)

@@

# set ggplot theme and colour palette
theme_set(theme_custom()) # set custom ggplot theme
clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient)

@@

## Word acquisition during toddlerhood: the effect of language experience

@@

Developmental psychology has increased the representativeness of its samples considerable in recent years, but most literature on early word acquisition relies extensively on data from monolingual children, and neglects to a considerable extent that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous research about bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals do. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology, as it is the case of English and Spanish. It is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

## Language distance as a mediator of bilingual vocabulary growth

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological similarity was measured by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. The resulting scores are then divided by the length of the longest string to normalise the similarity scores between 0 and 1, and then entered this variable as a predictor as they modelled participants' vocabulary sizes. Among other findings, the authors reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14% cognateness), were able to produce more Dutch words than English-Mandarin bilinguals (1.97% cognateness) were able to produce in Mandarin.

## Lexical non-selectiviy as a candidate account: the role of parallel activation

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results.  The parallel activation hypothesis stems from the language non-selective account of lexical access, and suggests that bilinguals activate both languages simultaneously during speech production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a plausible account for Floccia et al.'s results in the light of (1) the evidence pointing to bilingual toddlers' lexicon consisting of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage], and that (2) there is experimental evidence of parallel activation during word comprehension in children and toddlers [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. It is possible that cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is nonetheless scarce to date. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statiscally more reliable given their remarkably larger sample size. But although their results suggest a cognateness effect on vocabulary size, their study was not aimed at testing the effect of cognateness on age of acquisition at the *word-level.* The response variable in this study was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all words into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

In this study, we provide an account for the cognate facilitation on word acquisition rooted in accumulator models of language acquisition [@hidaka2013computational; @mollica2017how]. We build on the assumption that word acquisition is a function of mainly, but not exclusively, the amount of experience a child has accrued with the word-referent pairing [see @kachergis2022standard for review]. As infants accumulate linguistic input, they also accumulate opportunities to associate familiar word-forms to their corresponding referents. Words with high lexical frequency like *flower* [`r floor(childes[childes$token=="flower",]$freq_million)` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as *headphones* (`r floor(childes[childes$token=="headphones",]$freq_million)` times per million words) [e.g., @goodman2008does]: infants might accumulate learning experiences with *flower* faster than with *headphones*. In the case of bilinguals, the amount of exposures to a word-form an infants receives is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time might need longer to accumulate the same experience with *flower* and *headphones* than an English monolingual does (who listens 100% of the time to English).

We suggest that the cognateness effect proposed by @floccia2018results and @mitchell2022cognates facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *flower* and *flor*, for an English-Spanish bilingual) follow their acquisition trajectory in a similar way they would for a monolingual child: accumulating learning instances, which occur a a rate proportional to the lexical frequency of the word-form and the amount of exposure to the language they belong to. When one of the word-forms is acquired, the acquisition trajectory of the other is impacted, so that the number of needed learning instances left for acquisition are reduced. This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has acquired *flower*, and therefore its form-referent association has been formed, future presentations of the similar sounding translation *flor* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *dog*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014how; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*, which share four phonemes) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strenghtened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*, which do not share any phoneme) does not undergo the such facilitation. Both word-forms are fairly dissimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](img/hypotheses.jpg){#fig-hypotheses}

In line with this hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learnt the English word-form of translation equivalents earlier than the word-form in the additional language. Under this hypothesis, because of children's imbalance in exposure to words in English and in the additional language, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is unlikely to be available yet), while the acquisition of words in the additional language would benefit more frequently from their phonological similarity with the (more likely available) English translation.

## The present study

In this study, we tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a words *only* once its translation has been already acquired. Using an online vocabulary checklist designed specifically for this study, we collected vocabulary data on production and comprehension from a sample of bilingual children. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian item response theory [*IRT*, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word form, while adjusting for the the number of phonemes in the word form (an indicator of word acquisition difficulty).

We operationalised participants' exposure rate to a word-form using a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it when listening to speech), and  the child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with the participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childes] (see Appendix 1 for more details). Participant's degree of exposure to Catalan and Spanish were reported by families in a language exposure questionnaire they filled before the vocabulary checklist, in which they provided a proportion of the time that the child was estimated to listen to any language they were spoken to regularly. The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.

We defined *cognateness* as a continuous measure of phonological similarity between translation equivalents. We quantified the phonological similarity between each pair of word-forms by computing the inverse of the normalised Levenshtein distance between their SAMPA phonological transcriptions [@levenshtein1966binary]. This score is computed by counting the number of insertions, deletions and replacements needed by both transcriptions to become identical, dividing the resulting value by the length of the longest transcription, and finally subtracting this value from one. This results in a proportion that indicates how much the two phonological transcriptions of the translation equivalent are similar to each other, ranging from 0% (no similarity at all) to 100% (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. For instance, this measure of phonological similarity returns 0% for *ocell* (`uceL`) and *pájaro* (`paxa4o`), Catalan and Spanish for 'bird', and returns 50% for *lluna* (`Lun5`, /ˈʎu.nə/) and *luna* (`luna`, /ˈlu.na/), Catalan and Spanish for 'moon' (see Appendix 1 for more details).

>>>>

---
title: "Frequency of exposure mediates the facilitation effect of cognateness in bilingual lexical acquisition"

@@

abstract: |
    A prominent feature of the bilingual mental lexicon is that the activation of representations spreads in a language non-selective fashion. An instance of such non-selectivity is embodied by cognateness (i.e., form-similarity between translation equivalents, such as *flower*-*flor* in English and Spanish), which impacts lexical access and processing in bilinguals. Recent studies have suggested that cognateness facilitates vocabulary acquisition in bilingual toddlers---who show larger vocabulary sizes when their languages share many cognates---and an earlier age-of-acquisition for cognates than for non-cognates. The specific mechanisms underpinning such facilitation are unclear. We present an account of bilingual early lexical acquisition in which cognateness interacts with lexical frequency and language exposure to facilitate the acquisition of low-frequency words. We evaluated this model against vocabulary data from 436 Catalan-Spanish bilinguals aged 12 to 34 months. We used a Bayesian Exploratory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the toddler. We found converging evidence for an earlier age-of-acquisition for cognate words, and for such effect being mediated by lexical frequency and language exposure. Low-frequency words, and words from the language of least exposure were more strongly benefitted by their cognate status than high-frequency words. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

keywords: lexical acquisition, vocabulary, bilingualism, item response theory, bayesian
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(ggplot2)
library(patchwork)
library(bayesplot)
library(ggtext)
library(lubridate)
library(brms)
library(glue)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(scales)
library(bvqdev)

@@

tar_load(model_rhats)
tar_load(model_neffs)
tar_load(model_mcmc_pairs)
tar_load(model_ppcs)

@@

# set ggplot theme and colour palette
theme_set(theme_custom()) # set custom ggplot theme
bayesplot_theme_set(
    theme_custom()  +
        theme(panel.grid = element_blank(),
              axis.text = element_text(size = 5),
              strip.text = element_text(size = 5),
              plot.title = element_text(size = 8),
              plot.subtitle = element_text(size = 8))
)
clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")
color_scheme_set(rep(clrs[4], 6))
options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient)

@@

@@

Most literature on early word acquisition still relies extensively on data from monolingual children, and is oblivious to substantial proportion of the world population acquiring more than one language since early ages [@grosjean2021extent]. Previous research about bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology, as it is the case of English and Spanish. It is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological similarity was measured by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/kæt/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /æ/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /\textipa{kæt}/-/\textipa{Gata}/) would score three, as it requires at least three editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/ and adding another /a/. Levenshtein distance scores are then divided by the length of the longest string to normalise the similarity scores between 0 and 1. Following the previous examples: 1/3 (0.33) for /kæt/-/\textipa{kAt}/, and 3/4 (0.75) for /kæt/-/\textipa{Gata}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14% mean cognateness), were able to produce more Dutch words than English-Mandarin bilinguals (1.97% mean cognateness) were able to produce in Mandarin.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results.  The parallel activation hypothesis stems from the language non-selective account of lexical access, and suggests that bilinguals activate both languages simultaneously during speech production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain], even in fully monolingual environments. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a plausible account for Floccia et al.'s results in the light of (1) the evidence pointing to bilingual toddlers' lexicon consisting of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage], and that (2) there is experimental evidence of parallel activation during word comprehension in children and toddlers [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. It is possible that cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning languages with a larger proportion of cognates (such as children learning English and Dutch) showing larger vocabulary sizes. This account in which cognateness underlies the facilitation effect of typological language distance is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is scarce. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statistically more reliable given their remarkably larger sample size (372 participants). Although their results suggest a cognateness effect on vocabulary size, their study was not aimed at testing the effect of cognateness on age of acquisition at the *word-level.* The response variable in this study was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all words into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar. Word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

In the present investigation we approach the cognate facilitation on word learning using the conceptual framework of accumulator models of language acquisition [@hidaka2013computational; @mollica2017how]. We build on the assumption that word acquisition is a function of mainly---but not exclusively---the amount of experience a child has accrued with the corresponding word-referent pairing [see @kachergis2022standard for review]. As infants accumulate linguistic input, they also accumulate opportunities to associate familiar word-forms to their corresponding referents. Words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as 'headphones' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words) [e.g., @goodman2008does]. Infants might accumulate learning experiences with 'flower' faster than with 'taxi'. In the case of bilinguals, the amount of exposures to a word-form an infants receives is also mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual would (who listens 100% of the time to English).

We suggest that the cognateness effect proposed by @floccia2018results and @mitchell2022cognates facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *flower* and *flor*, for an English-Spanish bilingual) follow their acquisition trajectory in a similar way they would for a monolingual child: accumulating learning instances, which occur a a rate proportional to the lexical frequency of the word-form and the amount of exposure to the language they belong to. When one of the word-forms is sufficiently acquired, the acquisition trajectory of the other is impacted, so that the number of needed learning instances left for acquisition are reduced. This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has sufficient lexical knowledge of *flower*, and therefore there is some for-referent association in place, future presentations of the similar sounding translation *flor* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *dog*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*, which share four phonemes) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strengthened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*, which do not share any phoneme) does not undergo the such facilitation. Both word-forms are fairly dissimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](img/hypotheses.jpg){#fig-hypotheses}

In line with this hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learned the English word-form of translation equivalents earlier than the word-form in the additional language. Under this hypothesis, because of children's imbalance in exposure to words in English and in the additional language, the acquisition of English words by English-dominant participants would then benefit less frequently from their cognate status (the other word-form is unlikely to be available yet), while the acquisition of words in the additional language would benefit more frequently from their phonological similarity with the (more likely available) English translation.

We tested the hypothesis that the cognate effect on word acquisition is conditional to a child's exposure rate to the word, which would support the hypothesis that cognateness plays a role in the acquisition of a words *only* once its translation has been already acquired. We collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. Participants were aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months, and were exposed to Catalan and/or Spanish to varying degrees. We adopted a Bayesian explanatory item response theory  [*IRT*, see @kachergis2022standard for a similar approach] approach to model the probability of a given participant being reported to *understand* or *understand and say* each word-form in the checklist, conditional to participants age and rate of exposure to the word-form, and the cognate status of the word form, while adjusting for the the number of phonemes in the word form (an indicator of word acquisition difficulty).

We operationalised participants' exposure rate to a word-form using a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it), and child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childes] (see Appendix 1 for more details). Participants' degree of exposure were reported by caregivers in a language exposure questionnaire they filled before the vocabulary checklist. Caregivers provided a proportion of the time the child was estimated to listen to any language they were spoken to regularly (described in @sec-methods). The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.
